Effects of COVID-19 related social media use on well-being

### 2024-07-17 manuscript.Rmd
- updated before submission

Amidst the COVID-19 pandemic, staying informed became paramount, prompting heavy reliance on social media for updates, with use being at an all time high [@statistaAverageDailyTime2021]. 
The phenomenon of "doomscrolling" emerged as individuals struggled to disengage from COVID-19-related news [@buchananBriefExposureSocial2021; @sharmaDarkEndTunnel2022], sparking concerns about its impact on mental health [@sandstromDoomscrollingCOVIDNews2021]. 
While initial research began exploring this question [e.g., @bendauAssociationsCOVID19Related2021; @edenMediaCopingCOVID192020; @sewallObjectivelyMeasuredDigital2021], the extent of social media's influence on well-being remains largely unknown. 
This study aims to assess the effects of different social media usage patterns on individual well-being using a comprehensive longitudinal dataset spanning 34 waves, offering insights into within-person causal relationships.

@@

Social media is a broad term, encompassing subdimensions such as social networking sites (e.g., Facebook or Instagram), instant messengers (e.g., WhatsApp or Signal), or interactive video-platforms (e.g., YouTube or TikTok) [@carrSocialMediaDefining2015].
Effects of social media depend on how they are used: while purposeful, active, social use is generally related to positive outcomes, passive, non-purposeful and asocial use is related to negative outcomes [@frisonExploringRelationshipsDifferent2015; @verduynSocialNetworkingSites2022].
In this study, I hence distinguish three types of use and five popular channels. 
The types of use include posting (active use), reading (passive use), and liking and sharing (low-threshold active use) COVID-19 related content.
<!-- In doing so, this study analyzes social media use focused on COVID-19 related content. -->
<!-- , which includes posting thoughts about the pandemic, reading posts and comments , or liking and sharing COVID-19 related content. -->
The five channels to be investigated are Facebook, Twitter, Instagram, WhatsApp, and YouTube, which at the time ranked among the most popular social media services in Austria.

@@

<!-- Can social media use be such a factor? -->
Current literature overviews suggest that social media use on average does seem to decrease well-being [@meierComputermediatedCommunicationSocial2020a].
However, for most well-being outcomes, such as life satisfaction, general well-being, or loneliness, the effects are small [@meierComputermediatedCommunicationSocial2020a].
These small findings can be explained with the differential susceptibility of media effects model [@valkenburgDifferentialSusceptibilityMedia2013], which states that there is substantial _variation_ of media effects for individual users.
<!-- Whereas for some users social media are more beneficial, for others they are more harmful. -->
<!-- On average, however, effects tend to be small [@valkenburgDifferentialSusceptibilityMedia2013]. -->
For example, in one study it was estimated that roughly one quarter of all users experienced negative effects, another quarter positive effects, while for the rest the effects were neutral [@beyensSocialMediaUse2021].
<!-- Whether or not effects are positive or negative depend on (a) dispositional factors (e.g., personality, temperament, gender), (b) developmental factors (e.g., age, developmental tasks), (c) and social factors (e.g., environment, norms, upbringing) [@valkenburgDifferentialSusceptibilityMedia2013]. -->
<!-- Effects depend also on the content that is consumed---if the content is aligned with dispositions, developmental capacities, and converging contexts, effects tend to be stronger [@valkenburgDifferentialSusceptibilityMedia2013]. -->

Why are the effects of social media use on well-being small on average?
Two prominent media effect theories argue implicitly against strong average negative effects.
First, according to mood management theory [@zillmannMoodManagementCommunication1988], using media can affect people's moods.
<!-- Use can be stimulating or overwhelming, relaxing or boring. -->
After some time, users implicitly learn which media help them balance their mood and affect according to their own situational needs [@zillmannMoodManagementCommunication1988].
Those media that eventually become part of one's media repertoire hence, on average, tend to be beneficial for users to regulate their mood [@marcianoDynamicsAdolescentsSmartphone2022]. 
In conclusion, if a certain medium is used frequently, mood-management theory argues that it is likely not detrimental for well-being.

@@

On the other hand, the effects might be negative, perhaps best explained by the following five mechanisms.
Social media platforms can easily spread _false or misleading information_ about COVID-19 [@liYouTubeSourceInformation2020]. 
<!-- Due to the ease of sharing and the lack of fact-checking, inaccurate information can go viral and might cause confusion, anxiety, and panic among users. -->
Constant exposure to COVID-19-related content on social media can lead to _information overload_ and contribute to heightened anxiety levels [@fanInformationOverloadWellbeing2021]. 
<!-- The rapid spread of news, updates, and opinions can be overwhelming and might exacerbate existing stress or fears about the pandemic [@sharmaDarkEndTunnel2022]. -->
Discussions around COVID-19 are known for fostering negativity, with users sometimes engaging in _cyberbullying and harassment_. 
<!-- Discussions around COVID-19 can become heated and polarized, leading to personal attacks and online conflicts.  -->
<!-- Such experiences threaten mental well-being and might contribute to feelings of distress and isolation. -->
Social media often showcase the highlights and accomplishments of others, encouraging _social comparison_ [@przybylskiMotivationalEmotionalBehavioral2013]. 
During a pandemic, seeing posts about others' successes or seemingly perfect lives might intensify feelings of inadequacy or FOMO, especially when individuals are unable to participate in similar activities due to restrictions or personal circumstances [@sharmaDarkEndTunnel2022].

@@

These results align with prior research highlighting that social media usage is rather associated with elevated negative affect but not reduced life satisfaction [@meierComputermediatedCommunicationSocial2020a]. 
<!-- The often extreme, negative, or aggressive tone of COVID-19 discussions on social media [@fanStigmatizationSocialMedia2020] could adversely impact more active authors.  -->
<!-- The negative effects may be explained by the consumption of negative and misleading information [@liYouTubeSourceInformation2020]. -->
The study suggests that varying communication types and channels warrant separate analyses. 
Reading COVID-19 related content slightly reduced positive affect, while liking, sharing, and posting increased negative affect slightly. 
Posting COVID-19-related comments increased both negative and positive affect slightly, while it reduced life satisfaction. 
<!-- This suggests that posting generates stronger reactions, both positive and negative.  -->
Posting COVID-19 content was more negative for Generation Z, potentially reflecting broader negative effects of social media on this generation. 
<!-- Together, this finding is aligned with recent work challenging the assumption that active use is inherently beneficial and passive use detrimental [@valkenburgAssociationsActivePassive2022].  -->
Notably, communication channels exhibited differences. 
Twitter and YouTube appeared more negative, whereas Instagram, WhatsApp, and Facebook were neutral.

@@

Several limitations exist. While the study's focus on within-person effects enhances causal understanding, challenges exist related to additional relevant confounding exogenous variables not included here [@rohrerThinkingClearlyCorrelations2018], correct definition of the SESOI [@anvariUsingAnchorbasedMethods2021], and exact measurement of media use [@scharkowAccuracySelfreportedInternet2016]. 
The measures of social media use did not include interpersonal communication about COVID-19 related content, which might result in more positive outcomes [@frisonExploringRelationshipsDifferent2015].
The study's results are applicable primarily to Western societies and may not hold true in different cultural contexts.

>>>>

Amidst the COVID-19 pandemic, staying informed became paramount, prompting heavy reliance on social media for updates, with use being at an all time high [@statistaAverageDailyTime2021]. 
The phenomenon of "doomscrolling" emerged as individuals struggled to disengage from COVID-19-related news [@buchananBriefExposureSocial2021; @sharmaDarkEndTunnel2022], sparking concerns about its impact on mental health [@sandstromDoomscrollingCOVIDNews2021]. 
While initial research began exploring this question [e.g., @bendauAssociationsCOVID19Related2021; @edenMediaCopingCOVID192020; @sewallObjectivelyMeasuredDigital2021], the impact of COVID-19 related social media use on well-being remains largely unknown. 
This study aims to assess the effects of different social media usage patterns on individual well-being using a comprehensive longitudinal dataset spanning 34 waves, offering insights into within-person causal relationships.

@@

Social media is a broad term, encompassing subdimensions such as social networking sites (e.g., Facebook or Instagram), instant messengers (e.g., WhatsApp or Signal), or interactive video-platforms (e.g., YouTube or TikTok) [@carrSocialMediaDefining2015].
Effects of social media depend on how they are used: while purposeful, active, social use is generally related to positive outcomes, passive, non-purposeful and asocial use is related to negative outcomes [@frisonExploringRelationshipsDifferent2015; @verduynSocialNetworkingSites2022].
However, note that these general distinctions have been challenged recently [@meierDoesPassiveSocial2022, @valkenburgAssociationsActivePassive2022].
In this study, I hence distinguish three types of use and five popular channels. 
The types of use include posting (active use), reading (passive use), and liking and sharing (low-threshold active use) COVID-19 related content.
<!-- In doing so, this study analyzes social media use focused on COVID-19 related content. -->
<!-- , which includes posting thoughts about the pandemic, reading posts and comments , or liking and sharing COVID-19 related content. -->
The five channels to be investigated are Facebook, Twitter, Instagram, WhatsApp, and YouTube, which at the time ranked among the most popular social media services in Austria.

This study aims to provide a nuanced analysis of how a specific type of social media use influences various well-being outcomes. 
A general measure of time spent on social media would be insufficient to address how engagement with COVID-19-related content affects users. 
Conversely, using overly specific measures might be less broadly relevant or insightful. 
For example, demonstrating that receiving uplifting messages boosts social support or that encountering hate speech causes distress might yield limited insights. 
That said, ultimately, the specificity of a measure is not inherently right or wrong but reflects different approaches to answering different questions.

@@

<!-- Can social media use be such a factor? -->
Current literature overviews suggest that more active social media users report on average lower well-being [@meierComputermediatedCommunicationSocial2020a].
However, for most well-being outcomes, such as life satisfaction, general well-being, or loneliness, the effects are small [@meierComputermediatedCommunicationSocial2020a]. 
In addition, these results are based mostly on correlational research, and a recent meta study based on experimental studies did not find significant causal effects [@fergusonSocialMediaExperiments2024].
These small or even non-existent findings can be explained with the differential susceptibility of media effects model [@valkenburgDifferentialSusceptibilityMedia2013], which states that there is substantial _variation_ of media effects for individual users.
<!-- Whereas for some users social media are more beneficial, for others they are more harmful. -->
<!-- On average, however, effects tend to be small [@valkenburgDifferentialSusceptibilityMedia2013]. -->
For example, in one study it was estimated that roughly one quarter of all users experienced negative effects, another quarter positive effects, while for the rest the effects were neutral [@beyensSocialMediaUse2021].
<!-- Whether or not effects are positive or negative depend on (a) dispositional factors (e.g., personality, temperament, gender), (b) developmental factors (e.g., age, developmental tasks), (c) and social factors (e.g., environment, norms, upbringing) [@valkenburgDifferentialSusceptibilityMedia2013]. -->
<!-- Effects depend also on the content that is consumed---if the content is aligned with dispositions, developmental capacities, and converging contexts, effects tend to be stronger [@valkenburgDifferentialSusceptibilityMedia2013]. -->

Why are the effects of social media use on well-being small on average, especially given that several scholars theorize strong negative effects [@haidtAnxiousGenerationHow2024, @twengeIGENWhyToday2017]?
Two prominent media effect theories argue implicitly against strong average negative effects.
First, according to mood management theory [@zillmannMoodManagementCommunication1988], using media can affect people's moods.
<!-- Use can be stimulating or overwhelming, relaxing or boring. -->
After some time, users implicitly learn which media help them balance their mood and affect according to their own situational needs [@zillmannMoodManagementCommunication1988].
Those media that eventually become part of one's media repertoire hence, on average, tend to be beneficial for users to regulate their mood [@marcianoDynamicsAdolescentsSmartphone2022]. 
In conclusion, if a certain medium is used frequently, mood-management theory argues that it is likely not detrimental for well-being.

@@

On the other hand, the effects might be negative, perhaps best explained by the following five mechanisms.
Social media platforms can easily spread _false or misleading information_ about COVID-19 [@liYouTubeSourceInformation2020]. 
<!-- Due to the ease of sharing and the lack of fact-checking, inaccurate information can go viral and might cause confusion, anxiety, and panic among users. -->
Constant exposure to COVID-19-related content on social media can lead to _information overload_ and contribute to heightened anxiety levels [@fanInformationOverloadWellbeing2021]. 
<!-- The rapid spread of news, updates, and opinions can be overwhelming and might exacerbate existing stress or fears about the pandemic [@sharmaDarkEndTunnel2022]. -->
Discussions around COVID-19 are known for fostering negativity, with users sometimes engaging in _cyberbullying and harassment_. 
<!-- Discussions around COVID-19 can become heated and polarized, leading to personal attacks and online conflicts.  -->
<!-- Such experiences threaten mental well-being and might contribute to feelings of distress and isolation. -->
Social media often showcase the highlights and accomplishments of others, encouraging _social comparison_ [@przybylskiMotivationalEmotionalBehavioral2013]. 
Especially during a pandemic, seeing how others successfully cope with every-day challenges---for example, by baking banana bread, inventing creative games with kids, or exercising at home---might intensify feelings of inadequacy or FOMO, especially when individuals are unable to participate in similar activities due to restrictions or personal circumstances [@sharmaDarkEndTunnel2022].

@@

These results align with prior research highlighting that social media usage is rather associated with elevated negative affect but not reduced life satisfaction [@meierComputermediatedCommunicationSocial2020a]. 
<!-- The often extreme, negative, or aggressive tone of COVID-19 discussions on social media [@fanStigmatizationSocialMedia2020] could adversely impact more active authors.  -->
<!-- The negative effects may be explained by the consumption of negative and misleading information [@liYouTubeSourceInformation2020]. -->
The study suggests that varying communication types and channels warrant separate analyses. 
Reading COVID-19 related content slightly reduced positive affect, while liking, sharing, and posting increased negative affect slightly. 
Posting COVID-19-related comments increased both negative and positive affect slightly, while it reduced life satisfaction. 
<!-- This suggests that posting generates stronger reactions, both positive and negative.  -->
Posting COVID-19 content was more negative for Generation Z, potentially reflecting broader negative effects of social media on this generation. 
Together, this finding is aligned with recent work challenging the assumption that active use is inherently beneficial and passive use detrimental [@valkenburgAssociationsActivePassive2022].
Notably, communication channels exhibited differences. 
Twitter and YouTube appeared more negative, whereas Instagram, WhatsApp, and Facebook were neutral.

@@

Several limitations exist. While the study's focus on within-person effects enhances causal understanding, challenges exist related to additional relevant confounding exogenous variables not included here [@rohrerThinkingClearlyCorrelations2018], correct definition of the SESOI [@anvariUsingAnchorbasedMethods2021], and exact measurement of media use [@scharkowAccuracySelfreportedInternet2016]. 
As noted above, different operationalizations of COVID-19 related social media use might lead to different results. 
For example, the measures of social media use did not include interpersonal communication about COVID-19 related content, which might result in more positive outcomes [@frisonExploringRelationshipsDifferent2015].
While this study did not find meaningful within-person effects of social media use on well-being, between person analyses revealed strong negative correlations. 
Alternative causal processes not explored in this paper might explain these correlations.
The study's results are applicable primarily to Western societies and may not hold true in different cultural contexts.

### 2024-07-17 manuscript.tex
- updated before submission

\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Effects of COVID-19 related social media use on well-being},
  pdflang={en-EN},
  pdfkeywords={COVID-19, well-being, social media, news use, panel study.},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

@@

Amidst the COVID-19 pandemic, staying informed became paramount, prompting heavy reliance on social media for updates, with use being at an all time high (Statista, 2021).
The phenomenon of ``doomscrolling'' emerged as individuals struggled to disengage from COVID-19-related news (Buchanan et al., 2021; Sharma et al., 2022), sparking concerns about its impact on mental health (Sandstrom et al., 2021).
While initial research began exploring this question (e.g., Bendau et al., 2021; Eden et al., 2020; Sewall et al., 2021), the extent of social media's influence on well-being remains largely unknown.
This study aims to assess the effects of different social media usage patterns on individual well-being using a comprehensive longitudinal dataset spanning 34 waves, offering insights into within-person causal relationships.

@@

Social media is a broad term, encompassing subdimensions such as social networking sites (e.g., Facebook or Instagram), instant messengers (e.g., WhatsApp or Signal), or interactive video-platforms (e.g., YouTube or TikTok) (Carr \& Hayes, 2015).
Effects of social media depend on how they are used: while purposeful, active, social use is generally related to positive outcomes, passive, non-purposeful and asocial use is related to negative outcomes (Frison \& Eggermont, 2015; Verduyn et al., 2022).
In this study, I hence distinguish three types of use and five popular channels.
The types of use include posting (active use), reading (passive use), and liking and sharing (low-threshold active use) COVID-19 related content.
The five channels to be investigated are Facebook, Twitter, Instagram, WhatsApp, and YouTube, which at the time ranked among the most popular social media services in Austria.

@@

Current literature overviews suggest that social media use on average does seem to decrease well-being (Meier \& Reinecke, 2020).
However, for most well-being outcomes, such as life satisfaction, general well-being, or loneliness, the effects are small (Meier \& Reinecke, 2020).
These small findings can be explained with the differential susceptibility of media effects model (Valkenburg \& Peter, 2013), which states that there is substantial \emph{variation} of media effects for individual users.
For example, in one study it was estimated that roughly one quarter of all users experienced negative effects, another quarter positive effects, while for the rest the effects were neutral (Beyens et al., 2021).

Why are the effects of social media use on well-being small on average?
Two prominent media effect theories argue implicitly against strong average negative effects.
First, according to mood management theory (Zillmann, 1988), using media can affect people's moods.
After some time, users implicitly learn which media help them balance their mood and affect according to their own situational needs (Zillmann, 1988).
Those media that eventually become part of one's media repertoire hence, on average, tend to be beneficial for users to regulate their mood (Marciano et al., 2022).
In conclusion, if a certain medium is used frequently, mood-management theory argues that it is likely not detrimental for well-being.

@@

On the other hand, the effects might be negative, perhaps best explained by the following five mechanisms.
Social media platforms can easily spread \emph{false or misleading information} about COVID-19 (Li et al., 2020).
Constant exposure to COVID-19-related content on social media can lead to \emph{information overload} and contribute to heightened anxiety levels (Fan \& Smith, 2021).
Discussions around COVID-19 are known for fostering negativity, with users sometimes engaging in \emph{cyberbullying and harassment}.
Social media often showcase the highlights and accomplishments of others, encouraging \emph{social comparison} (Przybylski et al., 2013).
During a pandemic, seeing posts about others' successes or seemingly perfect lives might intensify feelings of inadequacy or FOMO, especially when individuals are unable to participate in similar activities due to restrictions or personal circumstances (Sharma et al., 2022).

@@

These results align with prior research highlighting that social media usage is rather associated with elevated negative affect but not reduced life satisfaction (Meier \& Reinecke, 2020).
The study suggests that varying communication types and channels warrant separate analyses.
Reading COVID-19 related content slightly reduced positive affect, while liking, sharing, and posting increased negative affect slightly.
Posting COVID-19-related comments increased both negative and positive affect slightly, while it reduced life satisfaction.
Posting COVID-19 content was more negative for Generation Z, potentially reflecting broader negative effects of social media on this generation.
Notably, communication channels exhibited differences.
Twitter and YouTube appeared more negative, whereas Instagram, WhatsApp, and Facebook were neutral.

Several limitations exist. While the study's focus on within-person effects enhances causal understanding, challenges exist related to additional relevant confounding exogenous variables not included here (Rohrer, 2018), correct definition of the SESOI (Anvari \& Lakens, 2021), and exact measurement of media use (Scharkow, 2016).
The measures of social media use did not include interpersonal communication about COVID-19 related content, which might result in more positive outcomes (Frison \& Eggermont, 2015).
The study's results are applicable primarily to Western societies and may not hold true in different cultural contexts.

@@

@@

@@

@@

@@

>>>>

\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Effects of COVID-19 related social media use on well-being},
  pdflang={en-EN},
  pdfkeywords={COVID-19, well-being, social media, news use, panel study.},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

@@

Amidst the COVID-19 pandemic, staying informed became paramount, prompting heavy reliance on social media for updates, with use being at an all time high (Statista, 2021).
The phenomenon of ``doomscrolling'' emerged as individuals struggled to disengage from COVID-19-related news (Buchanan et al., 2021; Sharma et al., 2022), sparking concerns about its impact on mental health (Sandstrom et al., 2021).
While initial research began exploring this question (e.g., Bendau et al., 2021; Eden et al., 2020; Sewall et al., 2021), the impact of COVID-19 related social media use on well-being remains largely unknown.
This study aims to assess the effects of different social media usage patterns on individual well-being using a comprehensive longitudinal dataset spanning 34 waves, offering insights into within-person causal relationships.

@@

Social media is a broad term, encompassing subdimensions such as social networking sites (e.g., Facebook or Instagram), instant messengers (e.g., WhatsApp or Signal), or interactive video-platforms (e.g., YouTube or TikTok) (Carr \& Hayes, 2015).
Effects of social media depend on how they are used: while purposeful, active, social use is generally related to positive outcomes, passive, non-purposeful and asocial use is related to negative outcomes (Frison \& Eggermont, 2015; Verduyn et al., 2022).
However, note that these general distinctions have been challenged recently Valkenburg et al. (2022).
In this study, I hence distinguish three types of use and five popular channels.
The types of use include posting (active use), reading (passive use), and liking and sharing (low-threshold active use) COVID-19 related content.
The five channels to be investigated are Facebook, Twitter, Instagram, WhatsApp, and YouTube, which at the time ranked among the most popular social media services in Austria.

This study aims to provide a nuanced analysis of how a specific type of social media use influences various well-being outcomes.
A general measure of time spent on social media would be insufficient to address how engagement with COVID-19-related content affects users.
Conversely, using overly specific measures might be less broadly relevant or insightful.
For example, demonstrating that receiving uplifting messages boosts social support or that encountering hate speech causes distress might yield limited insights.
That said, ultimately, the specificity of a measure is not inherently right or wrong but reflects different approaches to answering different questions.

@@

Current literature overviews suggest that more active social media users report on average lower well-being (Meier \& Reinecke, 2020).
However, for most well-being outcomes, such as life satisfaction, general well-being, or loneliness, the effects are small (Meier \& Reinecke, 2020).
In addition, these results are based mostly on correlational research, and a recent meta study based on experimental studies did not find significant causal effects (Ferguson, 2024).
These small or even non-existent findings can be explained with the differential susceptibility of media effects model (Valkenburg \& Peter, 2013), which states that there is substantial \emph{variation} of media effects for individual users.
For example, in one study it was estimated that roughly one quarter of all users experienced negative effects, another quarter positive effects, while for the rest the effects were neutral (Beyens et al., 2021).

Why are the effects of social media use on well-being small on average, especially given that several scholars theorize strong negative effects Twenge (2017)?
Two prominent media effect theories argue implicitly against strong average negative effects.
First, according to mood management theory (Zillmann, 1988), using media can affect people's moods.
After some time, users implicitly learn which media help them balance their mood and affect according to their own situational needs (Zillmann, 1988).
Those media that eventually become part of one's media repertoire hence, on average, tend to be beneficial for users to regulate their mood (Marciano et al., 2022).
In conclusion, if a certain medium is used frequently, mood-management theory argues that it is likely not detrimental for well-being.

@@

On the other hand, the effects might be negative, perhaps best explained by the following five mechanisms.
Social media platforms can easily spread \emph{false or misleading information} about COVID-19 (Li et al., 2020).
Constant exposure to COVID-19-related content on social media can lead to \emph{information overload} and contribute to heightened anxiety levels (Fan \& Smith, 2021).
Discussions around COVID-19 are known for fostering negativity, with users sometimes engaging in \emph{cyberbullying and harassment}.
Social media often showcase the highlights and accomplishments of others, encouraging \emph{social comparison} (Przybylski et al., 2013).
Especially during a pandemic, seeing how others successfully cope with every-day challenges---for example, by baking banana bread, inventing creative games with kids, or exercising at home---might intensify feelings of inadequacy or FOMO, especially when individuals are unable to participate in similar activities due to restrictions or personal circumstances (Sharma et al., 2022).

@@

These results align with prior research highlighting that social media usage is rather associated with elevated negative affect but not reduced life satisfaction (Meier \& Reinecke, 2020).
The study suggests that varying communication types and channels warrant separate analyses.
Reading COVID-19 related content slightly reduced positive affect, while liking, sharing, and posting increased negative affect slightly.
Posting COVID-19-related comments increased both negative and positive affect slightly, while it reduced life satisfaction.
Posting COVID-19 content was more negative for Generation Z, potentially reflecting broader negative effects of social media on this generation.
Together, this finding is aligned with recent work challenging the assumption that active use is inherently beneficial and passive use detrimental (Valkenburg et al., 2022).
Notably, communication channels exhibited differences.
Twitter and YouTube appeared more negative, whereas Instagram, WhatsApp, and Facebook were neutral.

Several limitations exist. While the study's focus on within-person effects enhances causal understanding, challenges exist related to additional relevant confounding exogenous variables not included here (Rohrer, 2018), correct definition of the SESOI (Anvari \& Lakens, 2021), and exact measurement of media use (Scharkow, 2016).
As noted above, different operationalizations of COVID-19 related social media use might lead to different results.
For example, the measures of social media use did not include interpersonal communication about COVID-19 related content, which might result in more positive outcomes (Frison \& Eggermont, 2015).
While this study did not find meaningful within-person effects of social media use on well-being, between person analyses revealed strong negative correlations.
Alternative causal processes not explored in this paper might explain these correlations.
The study's results are applicable primarily to Western societies and may not hold true in different cultural contexts.

@@

\bibitem[\citeproctext]{ref-fergusonSocialMediaExperiments2024}
Ferguson, C. J. (2024). Do social media experiments prove a link with mental health: {A} methodological and meta-analytic review. \emph{Psychology of Popular Media}. \url{https://doi.org/10.1037/ppm0000541}

@@

\bibitem[\citeproctext]{ref-haidtAnxiousGenerationHow2024}
Haidt, J. (2024). \emph{The anxious generation: How the great rewiring of childhood is causing an epidemic of mental illness}. Penguin Press.

@@

\bibitem[\citeproctext]{ref-meierDoesPassiveSocial2022}
Meier, A., \& Krause, H.-V. (2022). \emph{Does passive social media use harm well-being? {An} adversarial review} {[}Preprint{]}. PsyArXiv. \url{https://doi.org/10.31234/osf.io/nvbwh}

@@

\bibitem[\citeproctext]{ref-twengeIGENWhyToday2017}
Twenge, J. M. (2017). \emph{{iGEN}: {Why} today's super-connected kids are growing up less rebellious, more tolerant, less happy-and completely unprepared for adulthood}. Atria Books.

@@

\bibitem[\citeproctext]{ref-valkenburgAssociationsActivePassive2022}
Valkenburg, P. M., van Driel, I. I., \& Beyens, I. (2022). The associations of active and passive social media use with well-being: {A} critical scoping review. \emph{New Media \& Society}, \emph{24}(2), 530--549. \url{https://doi.org/10.1177/14614448211065425}
